### 2019-01-25 intro.tex
ssr

\subsection{Use of NCFs in Modeling Biological Systems}

The term \textbf{canalization} was first used by
Conrad Waddington (a British geneticist)
in 1942 \cite{Waddington-1942}
to express the ability of a species
to produce members with very similar observable features over a period
of time even when there are significant changes in the environment
during that period.
Since that time, the term ``canalization'' has been used to describe
the stability of a biological system with changes
in external conditions.
In 1969, Kauffman \cite{Kauffman-1969} introduced a Boolean network model
to explain the stability of gene regulatory networks.
Kauffman observed that the use of canalyzing Boolean
functions in the model captured many observed properties of
gene regulatory networks.
The subclass of NCFs was
introduced later by Kauffman et al. \cite{Kauffman-etal-2003} 
to facilitate a rigorous analysis of the Boolean network model
for gene regulatory networks.

@@

\subsection{Related Work}

Many researchers have studied mathematical properties of NCFs and
pointed out the importance of NCFs in modeling biological
phenomena (e.g., \cite{Kauffman-etal-2003,Kauffman-etal-2004,Layne-2011,
Layne-etal-2012,Li-etal-2011,Li-etal-2012,Li-etal-2013}).

\noindent
\textbf{Equivalence and Implication Problems.}~
We note that for general Boolean formulas, the implication and equivalence problems
are \cnp-hard \cite{GJ-1979}.

\noindent
\textbf{Sensitivity of Boolean Functions.}~
There is extensive literature on the sensitivity of various classes
of Boolean functions
(e.g., \cite{Buhrman-etal-2002,Nisan-etal-1994,Zhang-2011}).
For a discussion on how the sensitivity of a Boolean function provides
an indication of its stability, the reader is referred to \cite{Kauffman-etal-2004,
Layne-2011,Layne-etal-2012}.
Observations regarding relationships between sensitivities of Boolean functions
and their computational complexity are presented
in \cite{Buhrman-etal-2002,Nisan-etal-1994}.
Li and Adeyeye \cite{Li-etal-2012} present lower and upper bounds on the
sensitivity of any NCF.
Li et al. \cite{Li-etal-2011,Li-etal-2013} conjectured that the average
sensitivity of any NCF is strictly less than  $4/3$.
This conjecture was proved by Klotz et al. \cite{Klotz-etal-2013} by establishing
a tight upper bound on the average sensitivity of any NCF.
Their methods rely on Fourier analysis of Boolean functions
\cite{Odonnell-2014}.
In \cite{Stearns-etal-2018}, 
we presented a characterization of NCFs with maximum average sensitivity.
This characterization allowed us to present an alternative proof 
of the upper bound in \cite{Klotz-etal-2013} and derive a closed
form expression for the number of such NCFs.

>>>>

@@

\subsection{Related Work}
\label{sse:related}

The term \textbf{canalization} was first used by
Conrad Waddington (a British geneticist)
in 1942 \cite{Waddington-1942}
to express the ability of a species
to produce members with very similar observable features over a period
of time even when there are significant changes in the environment
during that period.
Since that time, the term ``canalization'' has been used to describe
the stability of a biological system with changes
in external conditions.
In 1969, Kauffman \cite{Kauffman-1969} introduced a Boolean network model
to explain the stability of gene regulatory networks.
Kauffman observed that the use of canalyzing Boolean
functions in the model captured many observed properties of
gene regulatory networks.
The subclass of NCFs was
introduced later by Kauffman et al. \cite{Kauffman-etal-2003} 
to facilitate a rigorous analysis of the Boolean network model
for gene regulatory networks.
Many researchers have studied mathematical properties of NCFs and
pointed out the importance of NCFs in modeling biological
phenomena (e.g., \cite{Layne-2011,
Layne-etal-2012,Li-etal-2011,Li-etal-2012,Li-etal-2013}).

There is extensive literature on the sensitivity of various classes
of Boolean functions
(e.g., \cite{Buhrman-etal-2002,Nisan-etal-1994,Zhang-2011}).
For a discussion on how the sensitivity of a Boolean function provides
an indication of its stability, the reader is referred to \cite{Kauffman-etal-2004,
Layne-2011,Layne-etal-2012}.
Li and Adeyeye \cite{Li-etal-2012} present lower and upper bounds on the
sensitivity of any NCF.
Li et al. \cite{Li-etal-2011,Li-etal-2013} conjectured that the average
sensitivity of any NCF is strictly less than  $4/3$.
This conjecture was proved by Klotz et al. \cite{Klotz-etal-2013} by establishing
a tight upper bound on the average sensitivity of any NCF.
Their methods rely on Fourier analysis of Boolean functions
\cite{Odonnell-2014}.
In \cite{Stearns-etal-2018}, 
we presented a characterization of NCFs with maximum average sensitivity.
This characterization allowed us to present an alternative proof 
of the upper bound in \cite{Klotz-etal-2013} and derive a closed
form expression for the number of such NCFs.

### 2019-01-25 preliminaries.tex
ssr

\subsection{Closure Properties of NCFs}

\begin{observation}\label{complement}
The set of NCFs and the set of generalized NCFs
are each closed under complement.
Moreover, given an NCF or a generalized NCF representation $f$, 
an NCF or generalized NCF $f'$ representing the complement of $f$ 
can be constructed in linear time.
\end{observation}

\noindent
\textbf{Proof:}~ We construct
representation $f'$ by complementing 
every canalyzed output in representation $f$.  \QED

\medskip
We also observe that the set of NCFs is not closed under OR,~ XOR~ or AND.
To see this for OR and XOR,  note that the functions 
$f_1 = x_1 \, \bar{x}_2$ and $f_2 = \bar{x}_1 \, x_2$
are each NCFs, but their OR and XOR is the same Boolean function,
and this Boolean function is not  a canalyzing function of either variable.
To see this for AND, note that the functions 
$f_1 = x_1 \vee \bar{x}_2$ and $f_2 = \bar{x}_1 \vee x_2$
are each NCFs, but their AND is not  a canalyzing function of either variable.
We now define a new operation (projection) on Boolean functions 
and show through an explicit construction that
NCFs are closed under this operation.
This construction is used in Section~\ref{sse:gen_ncf_implication}
where we present an efficient algorithm for the implication problem
for generalized NCFs.

\begin{definition}\label{def:projection}
Given a Boolean function $f$, a variable $x$, and a Boolean value $a$,
the {\bf projection} of $f$ on $x = a$,
denoted by $f_{x = a}$, is the function on the remaining variables
whose value on any assignment $\alpha$ to these variables
is the value of $f$ when $\alpha$ is extended to a complete assignment for $f$
by setting the variable $x$ to the value $a$.
\end{definition}

\noindent
We now show that generalized NCFs are closed under projection.

\begin{lemma}\label{lem:projection}
Given a generalized NCF representation $f$, a variable $x$, and a
Boolean value $a$, a generalized NCF representing the projection
$f_{x = a}$ can be constructed in linear time.
\end{lemma}

\noindent
\textbf{Proof:}~ We use a simple case analysis.
We will assume that the layers of the representation of $f$ have
been identified and that they are numbered 
from top to bottom, starting with 1.

\smallskip

\noindent
{\bf Case 1}:~ Variable $x$ does not occur in $f$.
Then  $f_{x = a}$ is the same as $f$.

\smallskip

\noindent
{\bf Case 2}:~ For some value $b$, $f$ contains the line~ 
``$x:~ a \longrightarrow b$".~
We have two subcases.
\begin{description}
\item {(i)} If this line occurs in the first layer of $f$, 
then $f_{x = a}$ is the constant $b$.

\item{(ii)} If this line occurs in a layer of $f$ other than the first layer, 
$f_{x = a}$ is obtained from $f$ by deleting 
the layer containing this line and all subsequent layers.
(Note that in this case, the preceding layer has canalyzed output $\bar{b}$.)
\end{description}

%%\smallskip

\noindent
{\bf Case 3}:~ For some value $b$, $f$ contains the line~
``$x:~ \bar{a} \longrightarrow b$".~
We have three subcases here.
\begin{description}
\item{(i)}
If this line occurs in a layer with more than one line, 
or if this line occurs in a layer that is not the last layer,
then $f_{x = a}$ is obtained from $f$ by deleting this line.
\item{(ii)}
If this line is the last line in $f$, 
and there is no other line in $f$ whose canalyzed output is $b$,
then $f_{x = a}$ is  the constant $\bar{b}$.
\item{(iii)}
If this line is the last line in $f$, is the only line is its layer, 
and there is at least one other line in $f$ whose canalyzed output is $b$,
then $f_{x = a}$ is obtained from $f$ by deleting the last two layers.
\end{description}
\QED

>>>>

\subsection{Defintions: Symmetric Boolean Functions}
\label{sse:symmetry}

A pair of variables of a Boolean function $f(x_1, x_2, \ldots, x_n)$ is
said to be \textbf{symmetric} if their values can be interchanged without
affecting the value of the function.
This form of interchange symmetry partitions the set of variables into a set of
 \textbf{symmetry groups}, where the members of each group are pairwise symmetric.
Function $f$ is said to be
$r$-\textbf{symmetric} if it has at most $r$ symmetry groups.
In this case, the value of $f$
depends only on how many of the variables in each symmetry group have the value 1.
We say that $f$ is \textbf{properly} $r$-\textbf{symmetric} if
it is $r$-symmetric, but not $(r-1)$-symmetric.

In the literature, a 1-symmetric function $f$ is referred to simply
as a \textbf{symmetric} function.
Thus, if $f$ is a symmetric function, then for any
input $(a_1, a_2, \ldots, a_n)$ to $f$, where $a_i \in \{0,1\}$ for
$1 \leq i \leq n$, and any permutation $\pi$ of $\{1, 2, \ldots, n\}$,
$f(a_1, a_2, \ldots, a_n)$ = $f(a_{\pi(1)}, a_{\pi(2)}, \ldots, a_{\pi(n)})$.

### 2019-01-25 symmetry_properties.tex
ssr

A pair of variables of a Boolean function $f(x_1, x_2, \ldots, x_n)$ is
said to be \textbf{symmetric} if their values can be interchanged without
affecting the value of the function.
This form of interchange symmetry partitions the set of variables into a set of 
 \textbf{symmetry groups}, where the members of each group are pairwise symmetric.
Function $f$ is said to be
$r$-\textbf{symmetric} if it has at most $r$ symmetry groups.
In this case, the value of $f$
depends only on how many of the variables in each symmetry group have the value 1.
We say that $f$ is \textbf{properly} $r$-\textbf{symmetric} if 
it is $r$-symmetric, but not $(r-1)$-symmetric.

In the literature, a 1-symmetric function $f$ is referred to simply 
as a \textbf{symmetric} function. 
Thus, if $f$ is a symmetric function, then for any
input $(a_1, a_2, \ldots, a_n)$ to $f$, where $a_i \in \{0,1\}$ for 
$1 \leq i \leq n$, and any permutation $\pi$ of $\{1, 2, \ldots, n\}$,
$f(a_1, a_2, \ldots, a_n)$ = $f(a_{\pi(1)}, a_{\pi(2)}, \ldots, a_{\pi(n)})$.

@@

@@

\begin{theorem}\label{thm:ncf_r_symmetric}
Suppose the normalized representation of a given NCF contains 
$r_1$ layers with only one distinct canalyzing value,
and $r_2$ layers with two distinct canalyzing values.
Then the function is properly $(r_1 + 2 r_2)$-symmetric.
\end{theorem}
\noindent
\textbf{Proof:}~
From Theorem \ref{thm:ncf_symmetric_variables}, 
any pair of variables occurring in the same layer, with the same canalyzing value,
are symmetric.
Thus the function is at most $(r_1 + 2 r_2)$-symmetric.
Moreover, any pair of variables from different layers, or with different canalyzing values
are not symmetric, so there are at least $(r_1 + 2 r_2)$ symmetry groups.
\QED

@@

A $r$-symmetric NCF has a normalized nested canalyzing representation with at most $r-1$ layers,
and has an unnormalized nested canalyzing representation with at most $r$ layers.
\end{corollary}

@@

{\bf Case 1:} $(x_{k'} : a_k \rightarrow b_k)$. \\
Since line $k'$ has the same canalyzing value and canalyzed output as line $k$,
and $f$ is properly $n$-symmetric,
line $k'$ must occur in a lower layer than that of line $k$.
Thus, there is at least one line between line $k$ and line $k'$ for which the canalyzed output is $\overline{b_k}$.
Let line $q$ be the first such line.
Note that for all $i$ such that $1 \leq i < q$, 
$c_{\pi(i)}$ does not match the canalyzing value of line $i$,
but $c_{\pi(q)}$ does match the canalyzing value of line $q$.
Since canalyzed output $b_q$ equals $\overline{b_k}$,
we have that $f(c_{\pi(1)}, c_{\pi(2)}, \ldots, c_{\pi(n)}) = \overline{b_k}$.

{\bf Case 2:} $(x_{k'} : a_k \rightarrow \overline{b_k})$. \\
Let line $q$ be the first line such that $k < q \leq k'$ and
the canalyzed output of line $q$ is $\overline{b_k}$.
Note that for all $i$ such that $1 \leq i < q$, 
$c_{\pi(i)}$ does not match the canalyzing value of line $i$,
so $f(c_{\pi(1)}, c_{\pi(2)}, \ldots, c_{\pi(n)}) = \overline{b_k}$.

{\bf Case 3:} $(x_{k'} : \overline{a_k} \rightarrow b_k)$. \\
Suppose there is a line below line $k$ for which the canalyzed output is $\overline{b_k}$.
Let $q$ be the first such line.
Note that for all $i$ such that $1 \leq i < q$, 
$c_{\pi(i)}$ does not match the canalyzing value of line $i$,
but $c_{\pi(q)}$ does match the canalyzing value of line $q$.
Thus, $f(c_{\pi(1)}, c_{\pi(2)}, \ldots, c_{\pi(n)}) = \overline{b_k}$.

@@

{\bf Case 4:} $(x_{k'} : \overline{a_k} \rightarrow \overline{b_k})$. \\
Suppose there is a line, other than line $k'$, below line $k$ for which the canalyzed output is $\overline{b_k}$.
Let $q$ be the first such line.
Note that for all $i$ such that $1 \leq i < q$, 
$c_{\pi(i)}$ does not match the canalyzing value of line $i$,
but $c_{\pi(q)}$ does match the canalyzing value of line $q$.
Thus, $f(c_{\pi(1)}, c_{\pi(2)}, \ldots, c_{\pi(n)}) = \overline{b_k}$.

@@

Suppose $f$ is $r$-symmetric for some $r < n$.
Then, by definition, there is a partition of the set $X$ =
$\{x_1, x_2, \ldots, x_n\}$ of variables into at most $r$ subsets such that
the value of $f$ depends only on the number of variables with value 1
in each subset.
Since $r < n$, this partition has a subset, say $X_j$, 
such that $|X_j| \geq 2$.
Let $X_j = \{x_{j_1}, x_{j_2}, \ldots, x_{j_k}\}$, where $k \geq 2$.
Assume without loss of generality that $j_1 < j_2 < \ldots < j_k$.
Define a permutation $\pi$ of $\{1, 2, \ldots, n\}$ as follows. 
\begin{enumerate}
\item For all $i$ such that $x_i \not\in X_j$, $\pi(i) = i$.
\item $\pi(j_{\ell})$ = $j_{\ell+1}$ for $1 \leq \ell_t \leq k-1$
      and $\pi(j_{k})$ = $j_1$. 
\end{enumerate}
Note that $\pi$ is not the identity permutation.
Thus, by Theorem~\ref{thm:ncf_asymmetry},
there is an input $(a_1, a_2, \ldots, a_n)$ to $f$ such that
such that
\begin{equation}\label{eqn:f_asymmetry}
f(a_1, a_2, \ldots, a_n) ~\neq~  f(a_{\pi(1)}, a_{\pi(2)}, \ldots, a_{\pi(n)})
\end{equation}
However, from the definition of permutation $\pi$, it follows
that the number of 1's in any subset of the partition 
is the same in both the inputs
$(a_1, a_2, \ldots, a_n)$ and
$(a_{\pi(1)}, a_{\pi(2)}, \ldots, a_{\pi(n)})$.
Therefore, since $f$ is $r$-symmetric, we must have
\begin{equation}\label{eqn:f_contra}
f(a_1, a_2, \ldots, a_n) ~=~ f(a_{\pi(1)}, a_{\pi(2)}, \ldots, a_{\pi(n)}).
\end{equation}
Equation~(\ref{eqn:f_contra}) contradicts 
Equation~(\ref{eqn:f_asymmetry}), and the result follows. \QED
\fi

>>>>

\subsection{Overview}
\label{sse:res_overview}

\subsection{Symmetric Canalyzing and Nested Canalyzing Functions}
\label{sse:sym_and_cf_ncf}

@@

\subsection{Symmetric Pairs of Variables in an NCF: A Characterization
and Its Applictions}
\label{sse:ncf_strong_sym}

@@

\begin{theorem}\label{thm:ncf_r_symmetric}
Suppose the normalized representation of a given NCF contains 
$r_1$ layers with only one distinct canalyzing value,
and $r_2$ layers with two distinct canalyzing values.
Then the function is properly $(r_1 + 2 r_2)$-symmetric.
\end{theorem}
\noindent
\textbf{Proof:}~
From Theorem~\ref{thm:ncf_symmetric_variables}, 
any pair of variables occurring in the same layer, with the same canalyzing value,
are symmetric.
Thus the function is at most $(r_1 + 2 r_2)$-symmetric.
Moreover, any pair of variables from different layers, or with different canalyzing values
are not symmetric, so there are at least $(r_1 + 2 r_2)$ symmetry groups.
\QED

@@

An $r$-symmetric NCF has a normalized nested canalyzing representation
with at most $r-1$ layers, and has an unnormalized nested canalyzing
representation with at most $r$ layers.  \end{corollary}

\subsection{Strong Asymmetry and NCFs}
\label{sse:strong_asym_ncf}

@@

\noindent
{\bf Case 1:} $(x_{k'} : a_k \rightarrow b_k)$. \\

Since line $k'$
has the same canalyzing value and canalyzed output as line $k$, and
$f$ is properly $n$-symmetric, line $k'$ must occur in a lower layer
than that of line $k$.  Thus, there is at least one line between
line $k$ and line $k'$ for which the canalyzed output is $\overline{b_k}$.
Let line $q$ be the first such line.  Note that for all $i$ such
that $1 \leq i < q$, $c_{\pi(i)}$ does not match the canalyzing
value of line $i$, but $c_{\pi(q)}$ does match the canalyzing value
of line $q$.  Since canalyzed output $b_q$ equals $\overline{b_k}$,
we have that $f(c_{\pi(1)}, c_{\pi(2)}, \ldots, c_{\pi(n)}) =
\overline{b_k}$.

\noindent
{\bf Case 2:} $(x_{k'} : a_k \rightarrow \overline{b_k})$. \\

Let line $q$ be the first line such that $k < q \leq k'$ and
the canalyzed output of line $q$ is $\overline{b_k}$.
Note that for all $i$ such that $1 \leq i < q$, 
$c_{\pi(i)}$ does not match the canalyzing value of line $i$,
so $f(c_{\pi(1)}, c_{\pi(2)}, \ldots, c_{\pi(n)}) = \overline{b_k}$.

\noindent
{\bf Case 3:} $(x_{k'} : \overline{a_k} \rightarrow b_k)$. \\

Suppose there is a line below line $k$ for which the canalyzed
output is $\overline{b_k}$.  Let $q$ be the first such line.  Note
that for all $i$ such that $1 \leq i < q$, $c_{\pi(i)}$ does not
match the canalyzing value of line $i$, but $c_{\pi(q)}$ does match
the canalyzing value of line $q$.  Thus, $f(c_{\pi(1)}, c_{\pi(2)},
\ldots, c_{\pi(n)}) = \overline{b_k}$.

@@

\noindent
{\bf Case 4:} $(x_{k'} : \overline{a_k} \rightarrow \overline{b_k})$. \\

Suppose there is a line, other than line $k'$, below line $k$ for
which the canalyzed output is $\overline{b_k}$.  Let $q$ be the
first such line.  Note that for all $i$ such that $1 \leq i < q$,
$c_{\pi(i)}$ does not match the canalyzing value of line $i$, but
$c_{\pi(q)}$ does match the canalyzing value of line $q$.  Thus,
$f(c_{\pi(1)}, c_{\pi(2)}, \ldots, c_{\pi(n)}) = \overline{b_k}$.

@@

Suppose $f$ is $r$-symmetric for some $r < n$.
Then, by definition, there is a partition of the set $X$ =
$\{x_1, x_2, \ldots, x_n\}$ of variables into at most $r$ subsets such that
the value of $f$ depends only on the number of variables with value 1
in each subset.
Since $r < n$, this partition has a subset, say $X_j$, 
such that $|X_j| \geq 2$.
Let $X_j = \{x_{j_1}, x_{j_2}, \ldots, x_{j_k}\}$, where $k \geq 2$.
Assume without loss of generality that $j_1 < j_2 < \ldots < j_k$.
Define a permutation $\pi$ of $\{1, 2, \ldots, n\}$ as follows. 
\begin{enumerate}
\item For all $i$ such that $x_i \not\in X_j$, $\pi(i) = i$.
\item $\pi(j_{\ell})$ = $j_{\ell+1}$ for $1 \leq \ell_t \leq k-1$
      and $\pi(j_{k})$ = $j_1$. 
\end{enumerate}
Note that $\pi$ is not the identity permutation.
Thus, by Theorem~\ref{thm:ncf_asymmetry},
there is an input $(a_1, a_2, \ldots, a_n)$ to $f$ such that
such that
\begin{equation}\label{eqn:f_asymmetry}
f(a_1, a_2, \ldots, a_n) ~\neq~  f(a_{\pi(1)}, a_{\pi(2)}, \ldots, a_{\pi(n)})
\end{equation}
However, from the definition of permutation $\pi$, it follows
that the number of 1's in any subset of the partition 
is the same in both the inputs
$(a_1, a_2, \ldots, a_n)$ and
$(a_{\pi(1)}, a_{\pi(2)}, \ldots, a_{\pi(n)})$.
Therefore, since $f$ is $r$-symmetric, we must have
\begin{equation}\label{eqn:f_contra}
f(a_1, a_2, \ldots, a_n) ~=~ f(a_{\pi(1)}, a_{\pi(2)}, \ldots, a_{\pi(n)}).
\end{equation}
Equation~(\ref{eqn:f_contra}) contradicts 
Equation~(\ref{eqn:f_asymmetry}), and the result follows. \QED
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\fi
